---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
echo = TRUE #make the code visible
unzip("activity.zip")
data <- read.csv("activity.csv") #load and read the data
head(data)
tail(data)
summary(data)
class(data)
str(data)

```

## What is mean total number of steps taken per day?
For this part of the assignment, I ignore the missing values in the dataset. 

### Calculate the total number of steps taken per day
```{r}
steps_date <- aggregate(steps ~ date, data, sum, na.action = na.omit)
head(steps_date)
tail(steps_date)
hist(steps_date$steps, main = paste("Total Steps Each Day"), col="steelblue", xlab="Number of Steps")
```

### Calculate and report the mean and median of the total number of steps taken per day
* The mean of the total number of steps taken per day: 
```{r}
mean(steps_date$steps)
```
* The median of the total number of steps taken per day: 
```{r}
median(steps_date$steps)
```


## What is the average daily activity pattern?

### a time series plot of the average number of steps taken versus the 5-minute intervals. 
```{r}
mean_steps_interval <- aggregate(steps ~ interval, data, mean,na.action = na.omit)
head(mean_steps_interval)

plot(mean_steps_interval$interval,mean_steps_interval$steps, type="l", col = "steelblue",lwd = 2,xlab="5-minute intervals", ylab="average number of steps taken",main="Average Number of Steps taken versus 5-minute intervals")
```

### 5-minute interval contains the maximum number of steps
```{r}
mean_steps_interval[mean_steps_interval$steps == max(mean_steps_interval$steps), ]
```



## Imputing missing values
```{r}
summary(data)
```
From the summry of the data, we can tell that there are 2304 NA's. 
We can also caculate it: 
```{r}
nrow(data[!complete.cases(data),])
```
### The strategy for filling all the missing values in the datasets 

Replacing the NA steps values with the average number of steps taken of the corresponding intervals as calculated above
```{r}
no_NA_data<-data
sapply(unique(data$interval),function(x)no_NA_data[!complete.cases(no_NA_data)&(no_NA_data$interval==x),1]<<-mean_steps_interval$steps[mean_steps_interval$interval==x])

```
no_NA_data is a new dataset that is euqal to the original data set but with the missing data filled in. 
```{r}
head(no_NA_data)
```
### A histogram of the total number of steps taken each day after missing values were imputed

```{r}
new_steps_date <- aggregate(steps ~ date, no_NA_data, sum)
hist(new_steps_date$steps, main = paste("Total Steps Each Day"), col="steelblue", xlab="Number of Steps")

```

* The mean of the total number of steps taken per day after missing data filled in: 
```{r}
mean(new_steps_date$steps)
```

* The median of the total number of steps taken per day after missing data filled in: 
```{r}
median(new_steps_date$steps)
```

* Calculate the difference before and after missing data filled in
```{r}
mean(steps_date$steps)-mean(new_steps_date$steps)
median(steps_date$steps)-median(new_steps_date$steps)
```
There is no difference for the mean, a slightly difference for the median.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
no_NA_data$date <- as.Date(no_NA_data$date, "%Y-%m-%d") ## change the date formate from factor to date
no_NA_data$work_or_not<-"weekday" ##assign everyday as weekday
no_NA_data$work_or_not[weekdays(no_NA_data$date)%in%c("Saturday","Sunday")]<-"weekend" ##assign days are weekends
no_NA_data$work_or_not<-as.factor(no_NA_data$work_or_not)
head(no_NA_data)
steps_by_interval<- aggregate(steps ~ interval + work_or_not, no_NA_data, mean)

library(lattice)

xyplot(steps_by_interval$steps ~ steps_by_interval$interval|steps_by_interval$work_or_not, main="Average Number of Steps taken versus 5-minute intervals",xlab="5-minute intervals", ylab="average number of steps taken",layout=c(1,2), type="l")

```
